In the Crucible: Massive Compact Galaxies with High-Velocity Outflows and Eddington-Limited Star Formation

### 2012-04-19 trunk/tex/sigmasfr/ms.tex
edits to section 3.2

Based on infrared photometry from the Wide-field Infrared Survey
Explorer (WISE) and optical morphologies from the Hubble Space
Telescope (HST), we present the discovery of compact, obscured star
formation in galaxies that exhibit $\gtrsim1000$~\kms\ outflows.  The
implied star formation rate (SFR) surface densities approach
$\sigmasfr=5000$~\msun~yr$^{-1}$~kpc$^{-2}$, and we argue that
feedback associated with this starburst event is a likely mechanism
for launching the high-velocity outflows.

@@

While the SEDs (Figure~\ref{fig:seds}) and optical spectra
(Figure~\ref{fig:spectra}) for our sample indicate that their
bolometric output is dominated by star formation, it is worthwhile to
consider the level of ongoing AGN activity and its effect on our
measurements.  Among the high-$\sigmasfr$ galaxies, the strongest case
for detectable AGN emission can be made for J1506+5402, which has the
most luminous \oiii~$\lambda5007$ line among the 29 galaxies observed
by HST.  This source also has a weak \nev~$\lambda3426$ emission line,
which is associated with AGN activity \citep[e.g.,][]{abe08,gil10}.
It was observed with the Chandra X-ray Observatory (proposal ID
11700896, PI: C. Tremonti) and had four detected counts, corresponding
to a 0.5--8.0~keV X-ray luminosity of $10^{42.7}$~erg~s$^{-1}$ (see
P. Sell et al., in preparation for more details on the 12/29 galaxies
with Chandra observations).  Using the relationship between 2--10~keV
X-ray and 12.3~$\mu$m mid-IR luminosity for AGNs from \citet{gan09},
one would expect a source with
$L_{\textnormal{\scriptsize{X}}}\approx5\times10^{42}$~erg~s$^{-1}$ to
have
$L_{\textnormal{\scriptsize{MIR}}}\approx7\times10^{42}$~erg~s$^{-1}$,
which is 400$\times$ smaller than observed luminosity for J1506+5402
($L_{\textnormal{\scriptsize{MIR}}}\approx3\times10^{45}$~erg~s$^{-1}$).
Based on its \oiii\ luminosity ($10^{42.1}$~erg~s$^{-1}$, a
significant fraction of which could be attributed to star formation)
and the calibration for type 1 AGNs from \citet{hec05}, one would
expect an intrinsic 2--10~keV luminosity of
$L_{\textnormal{\scriptsize{X}}}\approx5\times10^{43}$~erg~s$^{-1}$,
suggesting absorption by a factor of $\sim10$ in the X-rays.  However,
even if the X-ray attenuation were a factor of 100, which is the
typical value for local Compton-thick AGNs \citep{dia09}, the expected
mid-IR AGN contribution would only be relevant at the $\lesssim$30\%
level.  Similar to local ULIRGs \citep[e.g.,][]{far07}, the bolometric
output of the galaxies in our sample is dominated by star formation,
and our conclusions regarding large $\sigmasfr$ values are not
affected by AGN contamination.

@@

\subsection{Momentum Flux from Supernovae and Stellar Winds}

\citet{meu97} argued for a limit near 45~\units (somewhat smaller for
a Chabrier IMF) based on the surface brightness of starbursts measured
using rest-frame UV, H$\alpha$, far-IR, and radio continuum emission,
such that only 10\% of starbursts exceeded this limit.  They argued
that the momentum flux from supernovae and stellar winds expected at
this threshold matches the mean central pressure measured by
\citet{hec90} for starburst-driven winds.

There are also relevant arguments about ram pressure driving of
high-velocity winds from \citet{hec11}.

\subsection{The Eddington Limit from Radiation Pressure on Dust Grains}

In models where gaseous outflows are driven by radiation pressure from
massive stars, estimates of the maximum $\sigmasfr$ where the outward
force from radiation exceeds the inward force of gravity are in the
$\sigmasfr\sim10^3$~\units\ range \citep[e.g.,][]{tho05,hop10}.

@@

@@

>>>>

Based on infrared photometry from the Wide-field Infrared Survey
Explorer (WISE) and optical morphologies from the Hubble Space
Telescope (HST), we present the discovery of compact, obscured star
formation in galaxies that exhibit $\gtrsim1000$~\kms\ outflows.  The
implied star formation rate (SFR) surface densities approach
$\sigmasfr\approx5000$~\msun~yr$^{-1}$~kpc$^{-2}$, comparable to the
Eddington limit for a starburst where radiation pressure on dust
grains overcomes gravity.  We argue that feedback associated with this
starburst event, in the form of radiation pressure and supernova ram
pressure, is a likely mechanism for launching the high-velocity
outflows.

@@

While the SEDs (Figure~\ref{fig:seds}) and optical spectra
(Figure~\ref{fig:spectra}) for our sample indicate that their
bolometric output is dominated by star formation, it is worthwhile to
consider the level of ongoing AGN activity and its effect on our
measurements.  Among the high-$\sigmasfr$ galaxies, the strongest case
for detectable AGN emission can be made for J1506+5402, which has the
most luminous \oiii~$\lambda5007$ line among the 29 galaxies observed
by HST.  This source also has a weak \nev~$\lambda3426$ emission line,
which is associated with AGN activity \citep[e.g.,][]{abe08,gil10}.
It was observed with the Chandra X-ray Observatory (proposal ID
11700896, PI: C. Tremonti) and had four detected counts, corresponding
to a 0.5--8.0~keV X-ray luminosity of $10^{42.7}$~erg~s$^{-1}$ (see
P. Sell et al., in preparation for more details on the 12/29 galaxies
with Chandra observations).  Using the relationship between 2--10~keV
X-ray and 12.3~$\mu$m mid-IR luminosity for AGNs from \citet{gan09},
one would expect a source with
$L_{\textnormal{\scriptsize{X}}}\approx5\times10^{42}$~erg~s$^{-1}$ to
have
$L_{\textnormal{\scriptsize{MIR}}}\approx7\times10^{42}$~erg~s$^{-1}$,
which is 400$\times$ smaller than observed luminosity for J1506+5402
($L_{\textnormal{\scriptsize{MIR}}}\approx3\times10^{45}$~erg~s$^{-1}$).
Based on its \oiii\ luminosity ($10^{42.1}$~erg~s$^{-1}$, a
significant fraction of which could be attributed to star formation)
and the calibration for type 1 AGNs from \citet{hec05}, one would
expect an intrinsic 2--10~keV luminosity of
$L_{\textnormal{\scriptsize{X}}}\approx5\times10^{43}$~erg~s$^{-1}$,
suggesting absorption by a factor of $\sim10$ in the X-rays.  However,
even if the X-ray attenuation were a factor of 100, which is the
typical value for local Compton-thick AGNs \citep{dia09}, the expected
mid-IR AGN contribution would only be relevant at the $\lesssim$30\%
level.  Thus, similar to local ULIRGs \citep[e.g.,][]{far07}, the
bolometric output of the galaxies in our sample is dominated by star
formation, and we conclude that our results regarding large
$\sigmasfr$ values are not affected by AGN contamination.

@@

It is worth considering whether the high-velocity outflows we observe
were produced by the compact starburst.  \citet{mur11} argued that
massive star clusters with large gas surface densities are the ideal
launching point for galactic-scale outflows driven by radiation
pressure, and that the outflow velocity should scale with escape
velocity of the most massive star clusters in a galaxy.  The
$\gtrsim1000$~\kms\ outflow velocities we observe correspond
approximately to the escape velocities for a compact stellar
population with the sizes and masses that we've measured.
\begin{eqnarray}\nonumber
v_{esc}&=&\sqrt{2GM_{*}/r} \\
&=& 2100
\left({M_{*}\over10^{11} M_{\odot}}\right)^{1/2} 
\left({r\over200~\textnormal{pc}}\right)^{-1/2}
\textnormal{km~s}^{-1} 
\label{eq:vesc}
\end{eqnarray}
This argument, combined with the fact that we observe galaxies with
significant dust-obscured star formation and $\sigmasfr$ values near
the Eddington limit, indicates that momentum input from massive stars
in the form of radiation pressure is a viable mechanism for launching
these outflows.

In addition to radiation pressure, we also expect significant momentum
flux from stellar winds and supernovae.  For example, a starburst with
$\textnormal{SFR}\approx200$~\msun~yr$^{-1}$ would be associated with
radiation pressure $L_{bol}/c\approx2\times10^{35}$~dyne and ram
pressure $\dot{p}\approx5$--$10\times10^{35}$~dyne from stellar winds
and supernovae \citep[e.g.,][]{lei92,hec93,lei99,vei05}.  Furthermore,
\citet{hec11} noted that such a large momentum injection
($\dot{p}\approx10^{35}$~dyne) from a small initial radius
$r_{0}\approx100$~pc could accelerate a cloud with column density
$N_{H}\approx10^{21}$~cm$^{-2}$ to a terminal velocity
$v_{\infty}\approx1800$~\kms.  Thus, the energetics of compact
starbursts are sufficient to produce the high-velocity outflows we
observe, and it is plausible that both radiation pressure on dust
grains and supernova ram pressure contribute to driving the winds.

% Following \citet{hec93} and
% \citet{hec11}, the expected terminal velocity for a cloud with column
% density $\approx10^{21}$~cm$^{-2}$ launched from a radius
% $\approx200$~pc and driven by momentum flux
% $\approx2\times10^{35}$~dynes is:
% \begin{eqnarray}\nonumber
% v_{\infty}&=&1800~\textnormal{km~s}^{-1}
% \left(\dot{p}\over2\times10^{35}~\textnormal{dyne}\right)^{1/2} \\
% &&
% \left({r\over200~\textnormal{pc}}\right)^{-1/2}
% \left({N\over10^{21}~\textnormal{cm}^{-2}}\right)^{-1/2}
% \label{eq:vterm}
% \end{eqnarray}
% The expected momentum flux from stellar winds and supernovae for a
% starburst with $\textnormal{SFR}\approx200$~\msun~yr$^{-1}$ is
% $\approx5$--$10\times10^{35}$~dynes \citep{lei92,hec93,lei99,vei05}
% while the $L_{bol}/c$ radiation pressure is
% $\approx2\times10^{35}$~dynes.  Thus, there is more than enough
% momentum available from stellar winds and supernovae or from radiation
% pressure to produce the outflow velocities we observe.

%AGN have been invoked to launch high-velocity winds.

% \subsection{Momentum Flux from Supernovae and Stellar Winds}
% 
% \citet{meu97} argued for a limit near 45~\units (somewhat smaller for
% a Chabrier IMF) based on the surface brightness of starbursts measured
% using rest-frame UV, H$\alpha$, far-IR, and radio continuum emission,
% such that only 10\% of starbursts exceeded this limit.  They argued
% that the momentum flux from supernovae and stellar winds expected at
% this threshold matches the mean central pressure measured by
% \citet{hec90} for starburst-driven winds.
% 
% There are also relevant arguments about ram pressure driving of
% high-velocity winds from \citet{hec11}.
% 
% 
% \subsection{The Eddington Limit from Radiation Pressure on Dust Grains}

% In models where gaseous outflows are driven by radiation pressure from
% massive stars, estimates of the maximum $\sigmasfr$ where the outward
% force from radiation exceeds the inward force of gravity are in the
% $\sigmasfr\sim10^3$~\units\ range \citep[e.g.,][]{tho05,hop10}.

@@

Few galaxies exhibit such high-velocity outflows because few galaxies
have such extreme SFR surface densities.

@@

High-velocity outflows do not imply AGN feedback.

Theoretical arguments about ram v. radiation pressure.  Point that we
are observing 10$^4$~K gas.

Sample as a whole. Interesting that high-$\sigmasfr$ galaxies tend to
have larger outflow velocities.

Mention high-$\sigmasfr$ galaxies at high z.
